---
title: "Worked example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br>

### This worked example describes the use of *spmapper* and its core functions.

<br>

# Install *spmapper* 
You can install *spmapper* directly from within *R* (recommended) with the below code, or via download from the github repository, linked via the "Download spmapper" box on the homepage of this website. 
```{r eval=FALSE}
# Install spmapper from within R 
library(devtools)
devtools::install_github("NERC-CEH/spmapper-pkg") 

## If you have downloaded the package files directly from the github repository,  you will first need to install the package by modifying the below example code
# install.packages("C:/Users/myusername/Downloads/spmapper-pkg_[0.1.0].tar.gz", repos = NULL, type = "source") # with the correct file path and version name etc

# load the package
library(spmapper) 
```

```{r, echo=FALSE}
tooldir <- file.path(rprojroot::find_root(rprojroot::has_file("DESCRIPTION")), "inst/extdata")
```

# **Load packages and functions** 
Load packages needed for tool functions. Check you have all necessary packages installed. 

```{r, include=FALSE}
library(spmapper)
```

```{r message=FALSE, warning=FALSE}

# load required packages
library(terra) 
library(sf)
library(knitr)
library(rnaturalearth)
library(viridis)
library(ggplot2)
library(patchwork)

```
Define tool directory path - this is important to source the package functions, and for those functions to locate the input data!
```{r eval=FALSE}
## Define tool directory path
(tooldir <- system.file(package = "spmapper")) # automatically find and define your file path - but do check the path is correct

# tooldir <- file.path(".../spmapper") # alternatively, modify here to your package files directory if you have installed the package elsewhere
```

```{r, echo=FALSE}
tooldir <- system.file("extdata", package = "spmapper") # invisible code directory line
```

``` {r message=FALSE}
# Load package functions
invisible(system.file("R", "functions_preymodel.R", package = "spmapper"))
invisible(system.file("R", "functions_spmapper.R", package = "spmapper"))
invisible(system.file("R", "functions_fpudoverlap.R", package = "spmapper"))
invisible(system.file("R", "functions_spmapplot.R", package = "spmapper"))
```

<br><br>

# **Worked example** 

<br>

### **Specify user polygon(s)** 
Users can specify multiple polygon/multipolygon objects and apply the rest of the example code to compare prey consumption across areas. *spmapper* functions are capable of handling multipolygon objects. 
**Users are reminded to define the Coordinate Reference System (CRS) of their loaded shapefiles.** *spmapper* reprojects user polygons to align to the prey consumption maps. *spmapper* includes a function to plot prey consumption maps with the user polygons overlaid (see below).

```{r, eval=FALSE}
# Specify shapefile/s
fppolys <- sf::st_read(file.path()) # Add your filepath/s here. 

## Users must define the CRS of their loaded shapefile for spmapper to function!
fppolys <- st_crs([...]) # define source file CRS

```

<br>

**Example data presented**: UK offshore wind farm footprints of 'production', 'planned', 'construction' and 'approved' statuses, collated for the European Marine Observation and Data Network (via EMODnet; CETMAR, 2024). Offshore electricity transmission (OFTO) polygons excluded.
```{r echo=FALSE, error = FALSE, warning=FALSE, message=FALSE, fig.width = 5, fig.height = 5 , dpi = 500}
# Example polygons
# UK offshore windfarms (Sourced via EMODNet)
fppolys <- sf::st_read(file.path(tooldir,"demo/windfarmspolyPolygon.shp"), quiet = TRUE)
fppolys <- fppolys %>% dplyr::filter(country == "United Kingdom")
fppolys <- fppolys %>% dplyr::filter(status %in% c("Approved", "Construction", "Planned", "Production" )) 
fppolys <- fppolys %>% dplyr::filter(!grepl("OFTO", name)) # exclude cable polygons
proj.laea <- '+proj=laea +lat_0=-7.947 +lon_0=-14.30 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84  +units=m +no_defs'
fppolys <-  st_transform(fppolys, crs = proj.laea)

# Union of polygons to avoid double counting within overlapping polygons
fppolys <- st_union(fppolys) # combine overlapping elements 
fppolys <- st_sf(geometry = st_cast(fppolys, "POLYGON"))  # return to sfc


# UD example loaded as dummy data for polygon plot. UD not shown, prevented via fill scale in spmapplot. 
file_udmap <- system.file("extdata", paste0("FAME_UD_standardised_", "BLKI", ".tif"), package = "spmapper")
udmap <- terra::rast(file_udmap)
terra::crs(udmap) <- '+proj=laea +lat_0=-7.947 +lon_0=-14.30 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs +towgs84=0,0,0'
  
spmapplot(udmap, fppolys) + theme(legend.position = "none") # map colour scale wont work for UD values (used as dummy)

```

<br><br>

## **Run primary function *spmapper()* ** - produces prey consumption maps for all species and optionally extracts quantities within user-specified polygons, if provided. 

<br>

Optionally specify polygons with "fppolys = ". Users must specify the directory to tool data with "tooldir = " (defined earlier under 'Load packages and functions'). The object returned (assigned as the object named 'out' in this example) is a list object. This list contains 3 sub-lists (one per species) of the various inputs, quantities, and output prey consumption maps and extracted quantities. The species sub-lists are named by species, *e.g.* 'out\$Kittiwake'. The spmapper() output list also contains a dataframe summary of key results per species (named 'Summary', retrieved via 'out\$Summary'). The spmapper() function provides updates on its progress through the calculations. 


```{r error = FALSE, warning=FALSE, message=FALSE}
out <- spmapper(fppolys = fppolys, tooldir = tooldir)

```

<br>

Note, if you pass *spmapper()* a set of multiple polygons (*i.e.* a multipolygon object), spatial extractions will be performed and summed across all contained polygons. If you are interested in comparing two specific polygons - or two sets of polygons - you will need to run *spmapper()* separately for each, and then compare results. See example code below to run the tool for two individual or sets of input polygons, 'fppolys1' and 'fppolys2'. 

```{r eval=FALSE}
out1 <- spmapper(fppolys = fppolys1, tooldir = tooldir)
out2 <- spmapper(fppolys = fppolys2, tooldir = tooldir)
```

<br><br>

## **Visualise prey consumption maps with optional input polygons overlaid, using *spmapplot()* **

<br>

*spmapplot()* takes two arguments, the first is a prey consumption map (raster, generated by *spmapper()*, contained in list outputs in the form out\$Kittiwake\$*prey_cons_map*), the second optional argument is a polygon object (*e.g.* the user's polygons, *fppolys*)
```{r, , warning=FALSE,  message=FALSE, fig.width = 9, fig.height = 16, dpi = 500}

# Plot all species prey consumption maps 
(spmapplot(out$Kittiwake$prey_cons_map, fppolys) + ggtitle("Kittiwake")) +
(spmapplot(out$Guillemot$prey_cons_map, fppolys) + ggtitle("Guillemot")) +
(spmapplot(out$Razorbill$prey_cons_map, fppolys) + ggtitle("Razorbill")) + 
  patchwork::plot_layout(ncol = 1)
```

```{r, eval=FALSE}
# Or plot a single species map, e.g. Razorbill
spmapplot(out$Razorbill$prey_cons_map, fppolys) + ggtitle("Razorbill")

```


<br>

## **Visualise prey mass estimates**
The following code plots the distribution of prey mass estimates produced by *spmapper()*, and annotates the baseline estimate with a dashed line. 

```{r, fig.margin=10, , fig.width = 10, fig.height = 3.1, dpi = 500}

blki_p <- ggplot(data.frame(out$Kittiwake$allbirds_prey_sim), aes(out$Kittiwake$allbirds_prey_sim/1000)) +
  geom_histogram(bins = 50, color = "gray10", fill = "skyblue") +
  geom_vline(xintercept = out$Kittiwake$allbirds_prey_est/1000, color = "gray10", linetype = "longdash", linewidth = 1.1) +
  labs(x = "Prey consumption mass, t", y = "Frequency", title = "Kittiwake") +
  theme_minimal()

cogu_p <- ggplot(data.frame(out$Guillemot$allbirds_prey_sim), aes(out$Guillemot$allbirds_prey_sim/1000)) +
  geom_histogram(bins = 50, color = "gray10", fill = "skyblue") +
  geom_vline(xintercept = out$Guillemot$allbirds_prey_est/1000, color = "gray10", linetype = "longdash", linewidth = 1.1) +
  labs(x = "Prey consumption mass, t", y = "Frequency", title = "Guillemot") +
  theme_minimal()

razo_p <- ggplot(data.frame(out$Razorbill$allbirds_prey_sim), aes(out$Razorbill$allbirds_prey_sim/1000)) +
  geom_histogram(bins = 50, color = "gray10", fill = "skyblue") +
  geom_vline(xintercept = out$Razorbill$allbirds_prey_est/1000, color = "gray10", linetype = "longdash", linewidth = 1.1) +
  labs(x = "Prey consumption mass, t", y = "Frequency", title = "Razorbill") +
  theme_minimal()

(blki_p + cogu_p + razo_p) 

```

<br>

## **View key results**

The following code retrieves a table of  prey mass consumption (kg) estimates for each species and the results of the spatial extraction of prey consumption within polygons. Users can also access the full input/output data per species within the output sub-list for each (e.g. via 'out\$Kittiwake', and the entries therein).

*Estimate_kg* refers to the baseline *total* prey consumption (kg) estimate; *CI_Lower_kg* and *CI_Upper_kg* refer to the lower (2.5%) and upper (97.5%) confidence intervals, respectively.

*fp_percent* is the percentage of total prey consumption contained *within the polygons*. *fp_Estimate_kg* is the baseline prey mass (kg) contained within the polygons;	*fp_CI_Lower_kg* and *fp_CI_Upper_kg* refer to the lower (2.5%) and upper (97.5%) confidence intervals, respectively.



```{r eval = FALSE}

out$Summary

```

```{r echo = FALSE}
kable(out$Summary, caption = "spmapper Key Results")

```

<br><br><br>

___
# **Mis-specification of inputs**

*spmapper* performs checks for potential input specification errors. The following examples illustrate the error messages produced under cases of mis-specification.

<br>

### Incorrect path to tool
spmapper* does not currently explicitly check if the path provided is correct/valid, but may yield sensible error messages if an incorrect path is specified.

```{r error=TRUE}
spmapper(fppolys = fppolys, tooldir = "test")
```

<br>

## Validity of footprints
The tool does not check directly for the validity of the polygons provided, but does run three checks for consistency between polygons and the prey consumption maps. *spmapplot()* can be used to visualise polygons overlaid on the prey consumption maps.

<br>

### 1. Footprints on land (or in other areas of map without values) 

```{r echo=FALSE, error=TRUE, warning=FALSE, message=FALSE, fig.width = 6, out.width = '80%', dpi = 500}

# land polygon 
land_p <- st_sf(
  id = "land_p",
  geometry = st_sfc(
    st_polygon(list(rbind(
      c(715500 + 170000, 6651000),
      c(715500 + 170000, 6681000),
      c(745500 + 170000, 6681000),
      c(745500 + 170000, 6651000),
      c(715500 + 170000, 6651000)
    ))),
    crs = proj.laea
  )
)

spmapplot(out$Kittiwake$prey_cons_map, land_p)

spmapper(fppolys = land_p, tooldir = tooldir)

```

<br>

### 2. Footprints smaller than one grid cell 

```{r echo=FALSE, error=TRUE, warning=FALSE, message=FALSE, fig.width = 6, out.width = '80%', dpi = 500}

# 1.5 x 1.5 km polygon, i.e. too small for spmapper functionality
small_p <- st_sf(
  id = "small_p",
  geometry = st_sfc(
    st_polygon(list(rbind(
      c(1060250, 7011000),
      c(1060250, 7012500),
      c(1061500, 7012500),
      c(1061500, 7011000),
      c(1060250, 7011000)
    ))),
    crs = proj.laea
  )
)

tr_ext <- ext(725000, 1407000, 6653000, 7311000)

spmapplot(out$Kittiwake$prey_cons_map, small_p) + 
  coord_sf(
    xlim = c(xmin(tr_ext), xmax(tr_ext)),
    ylim = c(ymin(tr_ext), ymax(tr_ext)),
    expand = FALSE
  ) 

spmapper(fppolys = small_p, tooldir = tooldir)

```
<br>

### 3. Footprints outside extent of prey consumption maps

Note: this may be footprints on land, but it could also be footprints in areas that are within extent but in areas that weren't included in the original modelling in Wakefield et al. (2017). 

```{r echo=FALSE, error=TRUE, warning=FALSE, message=FALSE, fig.width = 6, out.width = '80%', dpi = 500}

outside_p <- st_sf(
  id = "outside_p",
  geometry = st_sfc(
    st_polygon(list(rbind(
      c(937000, 7391000),
      c(937000, 7451000),
      c(997000, 7451000),
      c(997000, 7391000),
      c(937000, 7391000)
    ))),
    crs = proj.laea
  )
)

ext_kitt <- ext(out$Kittiwake$prey_cons_map) 

spmapplot(out$Kittiwake$prey_cons_map, outside_p) + 
  geom_sf(data = outside_p,
          colour = "orchid2", fill = NA,
          alpha = 1, linewidth = 0.4) +
  coord_sf(xlim = c(138000, ext_kitt$xmax - 1000),
           ylim = c(5990000, 7485000),
           expand = FALSE)

spmapper(fppolys = outside_p, tooldir = tooldir)

```
<br><br>
